Crystal as a thermodynamic computer

### 2024-09-04 paper/main.tex

\begin{abstract}
\end{abstract}

@@

Intrigued by the question of how to thermalize and lower the energy of an Ising machine, we propose a new type of TC that has a crystalline structure.
The crystalline structure is based on a universal elementary cellular automaton, and is translational invariant in all spacial dimensions.
The translational invariance not only makes the crystal easier to be constructed in nature, but also enables the study of the scaling behavior the thermalization process.
In the rest of this paper, we first show that the crystalline structure can be programmed on its surfaces to perform universal computation \Cref{sec:Overview-of-main-results}.
Then we use numerical results in \Cref{sec:numerical-result} to show that if we thermalize the crystal from the so-called ``deterministic direction'', the crystal can thermalize in time almost linear to the size of the crystal, i.e. efficient computation can be performed.

@@

\section{Overview of the computation model}\label{sec:Overview-of-main-results}
In this article, We develop a new type of TC shown in \Cref{Surface-programmable-thermodynamic computer}(a), which features a crystalline structure based on a universal elementary cellular automaton, makes it easier to be constructed in real life.

@@

    \end{tikzpicture}
    \caption{Surface programmable thermodynamic computer (TC) and rule110 automaton gadget. $\mathtt{(a)}$ Example of a surface programmable TC with a moving heat bath slowly doing ``calculation''. White circles are spins and black lines represent interaction between spins, while red rectangles represent cellular automaton gadget. $\mathtt{(b)}$ Example of a rule110 gadget. Here spins $p$, $q$, and $r$ represent the left, middle, and right states in the cellular automaton, respectively, while $q’$ represents the result computed by the cellular automaton, and $A$ is an auxiliary spin. We define a spin pointing up as representing logical 1 and pointing down as representing logical 0. \jinguo{why capital A? q' not clear.}}
    \label{Surface-programmable-thermodynamic computer}
\end{figure}

@@

The input layer consists of the leftmost spin, while the output layer consists of the rightmost spin. The entire computing process can be described by gradually moving the external heat bath, which creates a temperature gradient and, consequently, a heat zone, from the input layer to the output layer. By utilizing thermal fluctuations from the heat bath, the spins in different layers sequentially settle into low-energy states within the state space, effectively completing the iterative computation of the cellular automaton.

@@

\subsection{Elementary cellular automaton}\label{elementary-cellular-automaton}

\jinguo{general CA, then the 1D CA.}
\ym{I haven't made change in this section}

An elementary cellular automaton is a 1-dimensional cellular
automaton, where there are two possible states (labeled by 0 and 1). The rule to
determine the state of the cell in next generation depends only on the
current state of the cell and its two immediate neighbors.

There are $8 = 2^3$ possible configurations for a cell and its two
immediate neighbors. Different elementary cellular automaton are only
different from their translation rules. There are only $2^8 = 256$
different rules, so do the automatons.

If we put each possible current configurations in order: 111, 110,
\ldots, 001, 000, and put the resulting state under them. We then get an
integer in its binary representations. Then this integer is taken to be
the rule number of the automaton. For example, rule 110.

Given that $110_d = 01101110_2$, so rule 110 is defined by the
translation rule:

@@

In boolean form, it can be expressed as:
\begin{equation}\label{eq:110rule}
    q^\prime = r \land \neg (q \land \neg p)
\end{equation}

Rule 110 has been shown to be Turing Complete~\cite{Cook2009}, and thus capable of universal
computation.

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (x \land y)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        H_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        H_{z = \neg c} &= s_c s_z\\
        H_{z = \neg (a \land b)} &= H_{z = \neg c} + H_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A cellular automaton can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the cellular automaton.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{}, we propose a provable optimal method to encode any cellular automaton rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}(b), which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}(b), respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
Since a copy gate\jinguo{do we really use copy gates?} \ym{No need in practice, but indeed used in Fig.1a} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming the TC shown in [Fig. \ref{Surface-programmable-thermodynamic computer}(a)]. The gadget we constructed based on the Rule 110 cellular automaton
naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

@@

The computational process of the cellular automaton is divided into three steps: determining the state of each input cell, iteratively applying the transition rules to generate the next generation of cells, and identifying a certain generation of cells as the output. In fact, the width of the time axis in the Ising TC corresponds to the number of iterations. Hence, the smallest layer of spins on the time axis should be the input layer, and the largest layer of spins should be the output layer.

Given that each ground state of the TC corresponds to a valid computational process of the cellular automaton for a given input, the computational process of the TC is essentially reduced to finding the ground state of the TC with the spin configuration in the input layers fixed.

@@

Given that simulating a cellular automaton with a deep computational layer on the Ising model requires a large number of spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.

Let us define a model Hamiltonian on a 2D lattice of size $S \times T$ as
\begin{equation}\label{eq:toy-hamiltonian}
H = \sum_{t=1}^{T-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}
where $n_{i, t} \in \{0, 1\}$ is the state of the cell at row $i$ and column $t$, the boundary is periodic in the space direction, i.e. $n_{0, t} = n_{S, t}$.
Its ground state with energy $0$ encodes a $110$ cellular automaton, where the direction of computation is from the left to right.
\subsection{Numerical result}

@@

We can encode the Rule 110 cellular automaton into a Weighted Maximum
Independent Set Problem, with blue vertices assigned a weight of 1 and
red vertices assigned a weight of 2, as follows.

@@

The states of vertex \textbf{1}, vertex \textbf{3}, and vertex
\textbf{8} represent the states of the \textbf{middle}, \textbf{left},
and \textbf{right} cells of the automaton's \textbf{input},
respectively. If the input value of a cell is 1, then the corresponding
vertex must be in the MWIS solution; otherwise, it is not. Vertex
\textbf{12} corresponds to the automaton's \textbf{output}. If the
automaton output is 1, then vertex 12 is in the MWIS solution;
otherwise, it is not.

In the automaton diagram, the above gadget is equivalent to:

@@

There are exactly \textbf{8} different MWIS solutions in this graph (the
weighted size of each MWIS solution is 7), each corresponding to one of
the \textbf{8} possible outputs of the automaton. We list them as
follows.
\includegraphics[width=\columnwidth]{../notes/images/gadget110.png}

@@

The above gadget depicts a two-layer cellular automaton. The vertices in
blue, red, green and black have weights of 1, 2, 3 and 4, respectively.
In the automaton diagram, the above gadget is equivalent to:

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../notes/images/rule110_2-2_automaton.png}
\caption{Alt text}
\end{figure}

>>>>

\begin{abstract}
    Thermodynamic computers are believed to be more energy efficient than conventional computers, however, a practical implementation of a thermodynamic computer is still challenging. In this paper, we propose a thermodynamics driven computer that enables use to perform computation by editing its surface.
    The crystalline structure is based on a universal elementary CA, and is translational invariant in all spacial dimensions. The translational invariance not only makes the crystal easier to be constructed in nature, but also enables the study of the scaling behavior the thermalization process.
\end{abstract}

@@

Intrigued by the question of how to thermalize and lower the energy of an Ising machine, we propose a new type of TC that has a crystalline structure.
The crystalline structure is based on a universal elementary CA, and is translational invariant in all spacial dimensions.
The translational invariance not only makes the crystal easier to be constructed in nature, but also enables the study of the scaling behavior the thermalization process.
In the rest of this paper, we first show that the crystalline structure can be programmed on its surfaces to perform universal computation \Cref{sec:Overview-of-main-results}.
Then we use numerical results in \Cref{sec:numerical-result} to show that if we thermalize the crystal from the so-called ``deterministic direction'', the crystal can thermalize in time almost linear to the size of the crystal, i.e. efficient computation can be performed.

@@

\section{Overview of the computation model}\label{sec:Overview-of-main-results}
In this article, We develop a new type of TC shown in \Cref{Surface-programmable-thermodynamic computer}(a), which features a crystalline structure based on a universal elementary CA, makes it easier to be constructed in real life.

@@

    \end{tikzpicture}
    \caption{Surface programmable thermodynamic computer (TC) and rule110 automata gadget. $\mathtt{(a)}$ Example of a surface programmable TC with a moving heat bath slowly doing ``calculation''. White circles are spins and black lines represent interaction between spins, while red rectangles represent CA gadget. $\mathtt{(b)}$ Example of a rule110 gadget. Here spins $p$, $q$, and $r$ represent the left, middle, and right states in the cellular automata, respectively, while $q^\prime$ represents the result computed by the cellular automata, and $A$ is an auxiliary spin. We define a spin pointing up as representing logical 1 and pointing down as representing logical 0. \jinguo{why capital A? q' not clear.}}
    \label{Surface-programmable-thermodynamic computer}
\end{figure}

@@

The input layer consists of the leftmost spin, while the output layer consists of the rightmost spin. The entire computing process can be described by gradually moving the external heat bath, which creates a temperature gradient and, consequently, a heat zone, from the input layer to the output layer. By utilizing thermal fluctuations from the heat bath, the spins in different layers sequentially settle into low-energy states within the state space, effectively completing the iterative computation of the CA.

@@

\subsection{Elementary cellular automata}\label{elementary-cellular-automata}
In cellular automata (CA), `space' is a regular lattice of `cells', each of which contains one of a small allowed set of integers.
Only cells that are close together interact in one `time-step'-the time evolution is given by a rule that looks at the contents of a few neighbouring cells, and decides what should change. At each step, this local rule is applied everywhere simultaneously~\cite{}.
Depending on different rules, some CAs can be more complex than others.
The most interesting category is those proved to be universal~\cite{Berlekamp2004}, meaning that they can simulate any computer program. The best-known examples in two-dimensional include Conway's~\cite{Games1970} ``Game of Life'', Margolus's reversible CA~\cite{Margolus1984} and its quantum version~\cite{Arrighi2012}.
The simplest universal CA is a 1-dimensional CA, namely, the rule 110 CA~\cite{Cook2009}.
This CA is an elementary CA that defined on a one dimensional grid, each cell has two possible states (labeled by 0 and 1). The rule of an elementary CA to determine the state of the cell in next generation depends only on the current state of the cell and its two immediate neighbors. The number of possible elementary CA rules is the same as the number of 3-input Boolean functions, which is $2^{2^3} = 256$. The Boolean function for rule 110 is shown in the following table:

@@

The name of the rule comes from the binary representation of the rule number, which is 110 in decimal.
Let us denote the states of the three cells as $p$, $q$, and $r$, from left to right, respectively. The rule 110 can be expressed as the following Boolean function:
\begin{equation}\label{eq:110rule}
    q^\prime = r \land \neg (q \land \neg p)
\end{equation}
where $q^\prime$ is the state of the center cell in the next generation.

@@

Previous studies have shown that the Ising model is universal for computation~\cite{Aadit2022, onizawa2020design, camsari2017stochastic}, meaning that any logic circuit can be encoded into the Ising model and solved by finding the ground state of the Ising model. A universal set logic gate gadgets are designed, by concatenating which any logic circuit can be constructed.
For example the logic expression $z = \neg (x \land y)$ can be encoded into the Ising model by the following Hamiltonian:
\begin{equation}
    \begin{split}
        H_{c = a \land b} &= s_a s_b -2 s_a s_c -2 s_bs_c + s_a + s_b -2 s_c\\
        H_{z = \neg c} &= s_c s_z\\
        H_{z = \neg (a \land b)} &= H_{z = \neg c} + H_{c = a \land b}
    \end{split}
\end{equation}
where, the variable $c$ is an ancilla spin.
A CA can be viewed as a translational invariant logic circuit, and thus can be encoded into the Ising model. The ground states of the Ising model encode the correspondence between the inputs and the outputs of the CA.
Directly synthesizing the logic circuit for the universal $110$ rule (\Cref{eq:110rule}) requires at least 3 ancilla spins, which is far from optimal. By converting the gadget finding problem to a linear programming problem~\cite{}, we propose a provable optimal method to encode any CA rule into the Ising model, which is described in \Cref{sec:gadget-design}.
The proposed gadget for the $110$ rule is shown in \Cref{Surface-programmable-thermodynamic computer}(b), which only requires 1 ancilla spin.
The interaction energy and onsite energy of the rule 110 gadget are given below, where $p, q, r$ and inputs and $q'$ is the output in \Cref{Surface-programmable-thermodynamic computer}(b), respectively, while spin $A$ is the auxiliary spin.
\begin{equation}
J = \begin{blockarray}{cccccc}
& p & q & r & q^\prime & A\\
\begin{block}{c(ccccc)}
p & ~\cdot~ & ~1~ & ~1~ & ~2~ & ~3~\\
q & \cdot & \cdot & 2 & 2 & 5\\
r & \cdot & \cdot & \cdot & 2 & 5\\
q^\prime & \cdot & \cdot & \cdot & \cdot & 6\\
A & \cdot & \cdot & \cdot & \cdot & \cdot\\
\end{block}
\end{blockarray}
~~, ~~h = \begin{blockarray}{cc}
& \phantom{x} \\
\begin{block}{c(c)}
p & ~1~\\
q & 2\\
r & 2\\
q^\prime & 2\\
A & 5\\
\end{block}
\end{blockarray}
\end{equation}
Since a copy gate\jinguo{do we really use copy gates?} \ym{No need in practice, but indeed used in Fig.1a} can be simply described as two spins with an interaction $J_{u, v}=1$, and by applying the method of connecting logic gates~\cite{Aadit2022}, we can concatenate the rule 110 gadget with some copy gates, ultimately forming the TC shown in [Fig. \ref{Surface-programmable-thermodynamic computer}(a)]. The gadget we constructed based on the Rule 110 CA
naturally possesses Turing completeness. Therefore, this 2-dimensional TC is capable of universal computing.

@@

The computational process of the CA is divided into three steps: determining the state of each input cell, iteratively applying the transition rules to generate the next generation of cells, and identifying a certain generation of cells as the output. In fact, the width of the time axis in the Ising TC corresponds to the number of iterations. Hence, the smallest layer of spins on the time axis should be the input layer, and the largest layer of spins should be the output layer.

Given that each ground state of the TC corresponds to a valid computational process of the CA for a given input, the computational process of the TC is essentially reduced to finding the ground state of the TC with the spin configuration in the input layers fixed.

@@

Given that simulating a CA with a deep computational layer on the Ising model requires a large number of spins, making large-scale numerical simulations impractical, we opted to introduce a similar but simpler energy model to observe the behavior on a larger scale.

Let us define a model Hamiltonian on a 2D lattice of size $S \times T$ as
\begin{equation}\label{eq:toy-hamiltonian}
H = \sum_{t=1}^{T-1}\sum_{i=1}^{S} (n_{i+1, t} \land \neg(n_{i, t} \land \neg n_{i-1, t})) \veebar n_{i,t+1},
\end{equation}
where $n_{i, t} \in \{0, 1\}$ is the state of the cell at row $i$ and column $t$, the boundary is periodic in the space direction, i.e. $n_{0, t} = n_{S, t}$.
Its ground state with energy $0$ encodes a $110$ CA, where the direction of computation is from the left to right.
\subsection{Numerical result}

@@

We can encode the Rule 110 CA into a Weighted Maximum
Independent Set Problem, with blue vertices assigned a weight of 1 and
red vertices assigned a weight of 2, as follows.

@@

The states of vertex \textbf{1}, vertex \textbf{3}, and vertex
\textbf{8} represent the states of the \textbf{middle}, \textbf{left},
and \textbf{right} cells of the automata's \textbf{input},
respectively. If the input value of a cell is 1, then the corresponding
vertex must be in the MWIS solution; otherwise, it is not. Vertex
\textbf{12} corresponds to the automata's \textbf{output}. If the
automata output is 1, then vertex 12 is in the MWIS solution;
otherwise, it is not.

In the automata diagram, the above gadget is equivalent to:

@@

There are exactly \textbf{8} different MWIS solutions in this graph (the
weighted size of each MWIS solution is 7), each corresponding to one of
the \textbf{8} possible outputs of the automata. We list them as
follows.
\includegraphics[width=\columnwidth]{../notes/images/gadget110.png}

@@

The above gadget depicts a two-layer CA. The vertices in
blue, red, green and black have weights of 1, 2, 3 and 4, respectively.
In the automata diagram, the above gadget is equivalent to:

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../notes/images/rule110_2-2_automata.png}
\caption{Alt text}
\end{figure}
